---
title: "Using GitHub Actions in R"
author: "Maximilian Held"
date: "`r Sys.Date()`"
output: rmarkdown::html_document
vignette: >
   %\VignetteIndexEntry{Rscript-byod}
   %\VignetteEngine{knitr::rmarkdown}
   %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Installation

If you haven't gotten access yet, [sign up](https://github.com/features/actions) for the closed beta of GitHub actions.

To install, run:

```r
devtools::install_github("maxheld83/ghactions")
```

Because you're likely only to ever use it *once*, **you need not take on ghactions as a dependency in your projects.**


## Quick Start

GitHub actions just requires a special file in a special directory at the root of your repository to work: `.github/main.workflow`.

To quickly set up such a file for frequently used project kinds, run:

```r
ghactions::use_ghactions(workflow = website())
```

(see the documentation for implied defaults).

Then push to GitHub and go to the actions tab in your repository.
Enjoy.

To learn more about GitHub actions and how to use it for and in R, read on.


## Docker First

Every GitHub *action* (not just the entire workflow) runs in its own little virtual machine, called a Docker container.
If you already know what this is, skip ahead.
If you're not familiar with Docker, it's easy to learn.

A Docker container exists in several forms.
It starts from a simple text file called `Dockerfile`, listing some instructions to build a virtual machine.
Think of this as the *recipe* to make your computing environment.
Using the `docker build` shell command, you can *build* a container *image*, which gets stored somewhere on your computer and which you can share with others.
Think of this as the prepared *meal*, which you've then put in the freezer for later.
Finally, you can `docker start` an image to boot said virtual machine.
Figuratively, you've now microwaved your pre-cooked meal and can *finally* enjoy it.

Docker images can also be *layered*.
For example, you can base a `Dockerfile` on an already existing docker *image*, and just add your *additional* instructions.
In the now tired frozen food analogy, you can use frozen pizza dough from the store as an *ingredient* to your own home-cooked meal.

Some GitHub actions use one of the many publicly available images to get their work done, such as the [action to publish to GitHub Pages](https://github.com/maxheld83/ghpages).
This would also make sense for those R projects, where you know *precisely* the build or runtime environment you need to adhere to, such as if you are deploying to [shinyapps.io](https://www.shinyapps.io) (their respective dockerfile is [here](https://github.com/rstudio/shinyapps-package-dependencies)).

For many other R projects, such as some R markdown project, there is *no* such standard computing environment, and you actually *want* to have the flexibility to provision one for yourself (say, because you need some obscure version of a system dependency such as the GSL or LaTeX).
In these situations, you have to **bring-your-own-`Dockerfile`**.
In fact, it's always safer to **byod**.
Happily, there are some great docker images for R projects maintained by the [Rocker Project](http://rocker-project.org).
Because containers can be layered, you can just base your *own* `Dockerfile` on one of those popular images. `rocker/verse` for example includes: R, RStudio, the tidyverse, LaTeX and many other packages and system dependencies.

A dockerfile is simply a text file called `Dockerfile` at the root of your repository.
At a minimum, it should include a `FROM` statement as in the below.

```
FROM rocker/verse:3.5.2
```
The `:3.5.2` part of the `FROM` statement binds you to a particular *version* of that image.
You are highly recommended to always use versioned images.
You can also generate such a simple file by running `ghactions::use_dockerfile()`.

If you want to add instructions, consult the [docker documentation](https://docs.docker.com/engine/reference/builder/).
Learn how to add additional R packages at the [Rocker Project](https://www.rocker-project.org/use/extending/).

You can also use the [containerit](https://github.com/o2r-project/containerit) package to automatically write `Dockerfile`: 
it uses impressive (and dark?) magic to figure out what dependencies you need.
Beginners might be better off writing their own `Dockerfile`s.

There is one last step before we can use the `Dockerfile` in a GitHub action.
Because a `Dockerfile` is just a *recipe* for a build environment, **you first have to build your `repo/Dockerfile`** into an image.
Happily, GitHub actions already provides an [action for that](https://github.com/actions/docker).
Most workflow function in this package start with a block such as the below.

```
action "Build image" {
  uses = "actions/docker/cli@c08a5fc9e0286844156fefff2c141072048141f6"
  args = "build --tag=repo:latest ."
}
```

This just runs `docker build --tag=repo:latest .`, where `.` is the root of your repository.
Docker will then pick whichever `Dockerfile` it finds there and *build* it.
Your dockerfile recipe has now been prepared into a meal, and this meal (the docker *image*) exists in your `/github/workspace`.
This is a special directory that you won't ever see anywhere.
GitHub actions provisions this directory and lets it *persist as long as your `main.workflow` runs*.

The `--tag=repo:latest` part of the above call simply names your image, literally "repo:latest".
This is just my convention, not a magic name.
The good news is that any downstream actions can now use the prepared image.

The awkward part is that these actions have to **call the image by its exact name: `repo:latest`**.
The workflow functions in this package take care of this automatically.
This isn't terribly elegant, but currently appears to be the only way on GitHub actions to identify images from past actions (see [this issue](https://github.com/maxheld83/ghactions/issues/1)).


## Actions

GitHub actions workflows are made up of individual blocks called -- you guessed it -- *actions*. 
In other CI/CD services, these are sometimes known as *steps*, though GitHub actions are more flexible.
You can read more about actions in [the GitHub documentation](https://developer.github.com/actions/).

This repository includes some custom R actions, but also relies greatly on outside actions (such as for deployment to, say, Netlify).
You can find a list in the actions category.

The below workflow-generating functions of this package add together individual actions and make them work together.

If you really wanted to, you can also use the templating functions to create action block syntax through ghactions:

```{r}
library(ghactions)
x <- rscript_byod(  # this just creates a list with sensible defaults
  IDENTIFIER = "Summation",
  needs = "Build image", # this is the necessary prior step
  # the uses field is hard-coded in the function to keep everything compatible
  fun = "1+1"
) 
do.call(what = make_action_block, args = x)
```

This isn't a very convincing use of R, and you'd be better off typing the action block by hand, or even just using the visual editor on GitHub.com.
To avoid duplication, there also isn't much documentation in these action wrappers; they just link to their respective repositories.
These action wrappers, in short, are useful only in the *very* narrow case where you might want to programmatically build your own templates.


## Workflows

This function generates `main.workflow` files for common workflow configurations of R projects.

Workflows are documented in the workflow category.


## Why You Should Care

It's still very early days, but it seems as if GitHub actions might have some potential for R projects:

- It's built on an open-source, cross-platform foundation, that you can easily reason about and reproduce on your machine (Docker!).
- It seems pretty darn fast compared to TravisCI: an `R CMD check` for a small project with all bells and whistles is at <4 minutes. (This is anecdotal evidence, if you'd like a real benchmarking, chip in [here](https://github.com/maxheld83/ghactions/issues/35)).
- There's one less service to deal with and one less authentication to go through, especially for smaller projects.
- The GUI makes it easier to build and reason about complex CI/CD chains.
- The sharing model seems promising.
- GitHub actions can be tied into almost *anything* in the GitHub API. Let's go nuts!

Of course there might be some vendor lock-in, though arguably less so than with other CI/CD services, because there's a thinner proprietary layer on top of Docker.


## What This Package *Doesn't* Do

The ghactions package is quite limited, and deliberately so:
GitHub actions already provides most of the things we might want, and in a cross-platform way:

- a succinct and human-readable text file representation (a subset of the [Hashicorp Configuration Language (HCL)](https://github.com/hashicorp/hcl)) of the [directed acyclic graphs (DAG)](https://en.wikipedia.org/wiki/Directed_acyclic_graph) used to model code automation workflows,
- a neat graphical user interface that makes to edit your `main.workflow` files in their native graph form,
- a convenient model and marketplace to share actions.

This package does not intend to solve this problem *again*, or to completely wrap GitHub actions in R.
It's really just a glorified collection of templates to get you started quickly.

If you need something more advanced, chances are you're going to want to edit your workflows yourself, using GitHub's native interface.
It's quite easy to use, and I'll try to gather and share best practices in this repository.


## Thanks

ghactions leaves much of the hard work to other open source software and their generous authors.

First and foremost, GitHub Actions is build on top of Docker, and so, by extension, is this package.
It would not work without the tremendous work of [Carl Boettiger](https://www.carlboettiger.info) and [Dirk Edelbuettel](http://dirk.eddelbuettel.com), who carefully maintain versioned Docker images for R through their [Rocker Project](http://rocker-project.org).

This package is also heavily modeled on, and indebted to the [usethis](https://usethis.r-lib.org) package by [Jenny Bryan](https://jennybryan.org) and [Hadley Wickham](http://hadley.nz).


## Related Work

There are plenty of other proven ways to run CI/CD for R.
Many rely on the [R support on TravisCI](https://docs.travis-ci.com/user/languages/r/), maintained by [Jeron Ooms](https://github.com/jeroen) and [Jim Hester](https://www.jimhester.com).
The [travis](https://ropenscilabs.github.io/travis/) and [tic](https://ropenscilabs.github.io/tic/) packages make it easier to work with them.
You can use [AppVeyor](http://appveyor.com)'s Windows-based system via the [r-appveyor](https://github.com/krlmlr/r-appveyor) package.

For serious, cross-platform testing of packages, there's the [r-hub](http://r-hub.io) project.

There are also other and additional ways to use R and Docker together.
The recommended way to use Dockere here is to simply edit a `DOCKERFILE` by hand, but the [containerit](http://o2r.info/containerit/) package can also, miraculously, do this for you by parsing your project files.
We are here primarily concerned with running R *inside Docker* (inside GitHub actions), but there are also some packages that allow you to control Docker from *inside R*, including [stevedore](https://richfitz.github.io/stevedore/), [harbor](https://github.com/wch/harbor) and [docker](https://bhaskarvk.github.io/docker/)., though it facilitates running Docker *inside R*, whereas GitHub Actions runs R *inside Docker*.
The broader topic of reproducibility in R with the help of Docker is also adressed by the [rrtools](https://github.com/benmarwick/rrtools) and [liftr](https://liftr.me) packages, as well as the [o2r](https://o2r.info) and [ropensci](https://ropensci.org) projects.
