---
title: "Running GitHub Actions Locally"
author: "Maximilian Held"
date: "`r Sys.Date()`"
output: rmarkdown::html_document
vignette: >
   %\VignetteIndexEntry{local}
   %\VignetteEngine{knitr::rmarkdown}
   %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r child='rmdhunks/disclaimer_docs.html'} 
```

For fast turnaround or debugging, you can also run GitHub actions locally on your computer.
For some projects, GitHub actions may even serve as a replacement for `make`.

Because it is based on Docker containers and an [open workflow format](https://github.blog/2019-02-07-an-open-source-parser-for-github-actions/), there are already several local, open-source workflow runners, including:

- [*act*](https://github.com/nektos/act)
- [*popper*](https://github.com/systemslab/popper)
- [*gha*](https://github.com/tschoffelen/gha)

None of these are officially supported by GitHub, and they may differ from the GitHub actions (the hosted service) in subtle ways.
Consult the respective issue [issue trackers](https://github.com/nektos/act/issues) for known limitations.

*act* currently appears to be the most comprehensive and mature alternative.
Installation and usage is documented [here](https://github.com/nektos/act) how to install it.


## Secrets

To run `act` with secrets, you can enter them interactively or supply them as environment variables.
If you have a secret called `FOO` in your `main.workflow`, `act` will take whatever you have set as `FOO` in the session from which you are running `act`.
If `FOO` is unset, it will ask you interactively.

You can set environment variables for the current session by running `export FOO="zap"`, or globally in your `.profile`.
You can also set environment variables *per directory* using a tool such as [direnv](https://direnv.net/).

**Be careful not to expose secrets**:
You may want to `.gitignore` any files or folders containing secrets, and/or encrypt secrets.


## Skipping Actions in `act`

You may sometimes want to skip some actions when you're running a `main.workflow` in act, such as deployment.
You can achieve something similar by using a [filter](https://github.com/actions/bin/tree/master/filter) action, filtering on all [`GITHUB_ACTOR`](https://developer.github.com/actions/creating-github-actions/accessing-the-runtime-environment/#environment-variables)s *except* `nektos/act`, which is the `GITHUB_ACTOR` set by `act`.

```
action "Filter Not Act" {
  uses = "actions/bin/filter@3c0b4f0e63ea54ea5df2914b4fabf383368cd0da"
  args = "not actor nektos/act"
}
```

Just remember that GitHub actions will cancel all upcoming and concurrent actions on a neutral exit code.
To avoid prematurely cancelling actions, place this filter at the latest possible point in the build graph.
